---
title: "Tutorial for main functions in survivalsurrogate package"
output:
  html_document:
    self_contained: true
---

```{r global_options, echo=FALSE}
knitr::opts_chunk$set(fig.width=7, warning=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=60))
```

If you are viewing the html version of this tutorial, you can find the R markdown file here: <https://github.com/denisagniel/survivalsurrogate>. It is the file called survivalsurrogate_tutorial.Rmd. In this tutorial, we will go through an example using the main functions in the **survivalsurrogate** package. First, install the package from CRAN and load it. We also load several other packages used in the document.

```{r results = "hide", message=FALSE}
#install.packages("survivalsurrogate")
library(survivalsurrogate)
library(ggplot2)
library(tidyr)
library(dplyr)
library(viridis) 
library(glue)
library(gbm)
library(mlr3)
library(rpart)
library(gbm)
library(mlr3extralearners)

```

This package provides influence function-based methods to evaluate a longitudinal surrogate marker in a censored time-to-event outcome setting, with plug-in and targeted maximum likelihood estimation options. More details will be available in the future in: Agniel D and Parast L (2025+). "Robust Evaluation of Longitudinal Surrogate Markers with Censored Data." Journal of the Royal Statistical Society: Series B, In press.

The functions in this package expect the data in a certain format. Specifically, **data** must be a dataframe with variables that indicate: (1) the folds for crossfitting; (2) a unique observation identifier; (3) baseline covariates, if there are none you must have a variable with all values equal to 1 to provide to the argument x below; (4) a variable indicating treatment group which should be 1 for treatment and 0 for control; (5) a set of variables that contain the surrogate marker value at each time point up to and including the landmark time, denoted t0; (6) a set of variables that indicate observation status at each time point where the surrogate marker is measured, in addition to the final time which is denoted as t; (7)  a set of variables that indicate primary outcome status at each time point where the surrogate marker is measured, in addition to the final time which is denoted as t. The **exampledata** in the package has the required format where t0=4 and t=5. Let's take a look. 

```{r}
data(exampledata)
names(exampledata)
```

In this dataset, **ID** is the observation ID, **X_0** is the single baseline covariate, **G_0** is the treatment indicator, and **ff** indicates the folds. The variables **Y** with a number are the variable indicate the primary outcome status at each time. For example, **Y_2** is 1 if the time of the primary outcome is greater than time 2. The variables **A** with a number indicate the observation status at each time. For example, **A_2** is 1 if the individual is still osbervable at time 2. The variables **S** with a number indicate the surrogate marker values at each time, up to and including t0=4. For example, **S_2** is the surrogate marker value at time 2. Notice that here, t0=4 and t=5 and thus, the dataset contains **Y_5** and **A_5** but does not contain **S_5**. 

Let's take a look at these surrogate marker measurements over time. If you don't care about these pictures, you can skip them. The first picture shows individual surrogate trajectories, and the second shows the same data but colored by treatment group.

```{r}
library(ggplot2)

# Pivot the data to long format
df_long <- exampledata %>%
  mutate(ID = as.factor(ID)) %>%  # ensure ID is a factor
  pivot_longer(
    cols = starts_with("S_"),
    names_to = "Time",
    names_prefix = "S_",
    values_to = "S_value"
  ) %>%
  mutate(
    Time = as.numeric(Time)
  )


#surrogate marker values over time
ggplot(df_long, aes(x = Time, y = S_value, group = ID, color = ID)) +
  geom_line(alpha = 0.8, linewidth = 1, na.rm = TRUE) +
  scale_color_viridis_d(option = "rocket") +
  labs(title = "Individual Surrogate Trajectories",
       x = "Time",
       y = "Surrogate Value") +
  theme_minimal() +
  theme(legend.position = "none")

#now color by treatment group
ggplot(df_long, aes(x = Time, y = S_value, group = ID, color = factor(G_0))) +
  geom_line(alpha = 0.6, linewidth = 1, na.rm = TRUE) +
  scale_color_manual(values = c("pink", "purple"), labels = c("Control", "Treatment")) +
  labs(title = "Surrogate Trajectories by Treatment Group",
       x = "Time",
       y = "Surrogate Value",
       color = "Treatment") +
  theme_minimal()


```

The goal is to estimate the proportion of the treatment effect on the primary outcome at time t=5 that is explained by the longitudinal surrogate marker up to t0=4. Values close to 1 indicate a strong surrogate whereas values close to 0 indicate a weak surrogate. The package offers estimation via a plug-in estimator or targeted maximium likelihood estimation (TMLE). Let's start with the plug-in estimator. In the code below, we set up t as tt=5 and t0 as t0=4; we set the learner to be gbm (generalized boosted model); and specify which variables indicate the **Y**, **A**, and **S**. We first estimate the treatment effect on the primary outcome (delta) using **plugin_delta**, and then the residual treatment effect (delta_s) using **plugin_delta_s**. Next, we give these results to the **estimate_R** function to calculate the proportion. The default method for variance estimation is "asymptotic" but you can change this to the bootstrap by specifying **se_type = "asymptotic"** in the arguments; if you do this, you will need to specify a number for the bootstrap sample, for example **n_boot=200**.

```{r}
tt <- 5
t0 <- 4
yvars <- paste0('Y_', 0:tt)
lrn <- 'gbm'
lrnc <- glue('regr.{lrn}')
lrnb <- glue('classif.{lrn}')

p_deltahat <- plugin_delta(
    data = exampledata,
    folds = 'ff',
    id = 'ID',
    x = 'X_0',
    g = 'G_0',
    a = paste0('A_', 0:tt),
    y = yvars,
    s = paste0('S_', 0:t0),
    binary_lrnr = lrn(lrnb, predict_type = 'prob'),
    cont_lrnr = lrn(lrnc)
  )

p_deltahat_s <- plugin_delta_s(
    data = exampledata,
    folds = 'ff',
    id = 'ID',
    x = 'X_0',
    g = 'G_0',
    a = paste0('A_', 0:tt),
    y = yvars,
    s = paste0('S_', 0:t0),
    binary_lrnr = lrn(lrnb, predict_type = 'prob'),
    cont_lrnr = lrn(lrnc),
    t0=t0
  )

                                 
estimate_R(p_deltahat$if_data[[1]]$eif,
                       p_deltahat_s$if_data[[1]]$eif) 

```
The output shows the estimate, standard error (se), and lower and upper bounds for the 95\% confidence interval (ci_l and ci_h), for each estimand, where **R** is the proportion of the treatment effect explained.

Now we use TMLE using the functions **tmle_delta** and **tmle_delta_s**:

```{r}
tml_deltahat <- tmle_delta(data = exampledata,
                             folds = 'ff',
                             id = 'ID',
                             x = 'X_0',
                             g = 'G_0',
                             a = paste0('A_', 0:tt),
                             y = yvars,
                             s = paste0('S_', 0:t0),
                             binary_lrnr = lrn(lrnb, predict_type = 'prob'),
                             cont_lrnr = lrn(lrnc))

tml_deltahat_s <- tmle_delta_s(data = exampledata,
                                 folds = 'ff',
                                 id = 'ID',
                                 x = 'X_0',
                                 g = 'G_0',
                                 a = paste0('A_', 0:tt),
                                 y = yvars,
                                 s = paste0('S_', 0:t0),
                                 binary_lrnr = lrn(lrnb, predict_type = 'prob'),
                                 cont_lrnr = lrn(lrnc))

                                 
estimate_R(tml_deltahat$if_data[[1]]$eif,
                       tml_deltahat_s$if_data[[1]]$eif) 

```

Again, the output shows the estimate, standard error (se), and lower and upper bounds for the 95\% confidence interval (ci_l and ci_h), for each estimand, where **R** is the proportion of the treatment effect explained.


That's all for now!

---------